---
title: "rsleep: A R package for sleep data analysis"
pagetitle: rsleep, A R package for sleep data analysis
output: github_document
always_allow_html: yes
bibliography: README.bibtex
csl: README.csl
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

[![CRAN](http://www.r-pkg.org/badges/version/rsleep)](https://cran.r-project.org/package=rsleep) [![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/rsleep)](https://cran.r-project.org/package=rsleep) [![Build Status](https://travis-ci.org/boupetch/rsleep.svg?branch=master)](https://travis-ci.org/boupetch/rsleep) [![codecov](https://codecov.io/gh/boupetch/rsleep/branch/master/graph/badge.svg)](https://codecov.io/gh/boupetch/rsleep) [![License:MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT) 

rsleep is a R package providing tools for sleep data management, visualization and analysis. 

## Installation

Stable version can be downloaded and installed from [CRAN](https://cran.r-project.org/) @hornik2012comprehensive :

```{r install_cran, results="hide"}
install.packages("rsleep", repos = "https://cloud.r-project.org")
```

Development version can be directly installed from [Github](https://github.com/) using the `devtools` package :

```{r install_github, results="hide", message=FALSE}
devtools::install_github("boupetch/rsleep")
```

## Usage

```{r library,results="hide"}
library(rsleep)
```

## Vignettes

  * [Managing Sleep Records Files](http://htmlpreview.github.io/?https://raw.githubusercontent.com/boupetch/rsleep/master/doc/Managing_Sleep_Records_Files.html)
  * [Spectral analysis of sleep electroencephalography signals](http://htmlpreview.github.io/?https://raw.githubusercontent.com/boupetch/rsleep/master/doc/Spectral_analysis_sleep_electroencephalography.html)
  * [Automatic Stages Classification](http://htmlpreview.github.io/?https://raw.githubusercontent.com/boupetch/rsleep/master/doc/Automatic_Stage_Classification.html)

## References
