---
title: "rsleep: A R package for sleep data analysis"
pagetitle: rsleep, A R package for sleep data analysis
output: github_document
always_allow_html: yes
bibliography: README.bibtex
csl: README.csl
editor_options: 
  chunk_output_type: console
---

[![CRAN](http://www.r-pkg.org/badges/version/rsleep)](https://cran.r-project.org/package=rsleep) [![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/rsleep)](https://cran.r-project.org/package=rsleep) [![Build Status](https://travis-ci.org/boupetch/rsleep.svg?branch=master)](https://travis-ci.org/boupetch/rsleep) [![codecov](https://codecov.io/gh/boupetch/rsleep/branch/master/graph/badge.svg)](https://codecov.io/gh/boupetch/rsleep) [![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT) 

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

## Installation

`rsleep` development version can be directly installed from Github using the `devtools` package.

```
devtools::install_github("boupetch/rsleep")
```

Otherwise stable version can be downloaded and installed from CRAN @hornik2012comprehensive :

```
install.packages("rsleep")
```

```{r library, include=FALSE}

library(rsleep)

```

## Getting sleep data

An example sleep record can be downloaded using the following code line. It contains full polysomnography data recorded over a whole night. Signals form sensors are stored in the European Data Format @kempSimpleFormatExchange1992 (EDF) file, while events are stored in the Comma-Separated Values (CSV) file,

```{r download_data_hidden, include=FALSE}

if(!file.exists("15012016HD.edf")){
  download.file("http://cloud.frenchkpi.com/s/65cm6DMq7SYKQ6J/download", "15012016HD.edf")
}

if(!file.exists("15012016HD.csv")){
  download.file("http://cloud.frenchkpi.com/s/wreGqkitWNnWwnP/download", "15012016HD.csv")
}

```


```{r download_data_display, eval=FALSE}

download.file("http://cloud.frenchkpi.com/s/65cm6DMq7SYKQ6J/download", "15012016HD.edf")

download.file("http://cloud.frenchkpi.com/s/wreGqkitWNnWwnP/download", "15012016HD.csv")

```

## Records manipulation

In `rsleep`, `write_mdf()` and `read_mdf()` functions are used to write and read records on disk. Files are converted from the EDF to Morpheo Data Format @bouchequetMorpheoDataFormat2018a (MDF). MDF is a simple, efficient and interoperable hierarchical file format for biological timeseries. The format supports raw signal and metadata storage. MDF uses binary files for signals and JSON for metadata.

```{r write_mdf_hidden, eval=FALSE}

if(!dir.exists("15012016HD")){
  events <- read_events_noxturnal("15012016HD.csv")

  write_mdf(edfPath = "15012016HD.edf",
            mdfPath = "15012016HD",
            channels = c("C3-M2", "ECG"),
            events = events)
}

```

```{r write_mdf_display, eval=FALSE}

events <- read_events_noxturnal("15012016HD.csv")

write_mdf(edfPath = "15012016HD.edf",
          mdfPath = "15012016HD",
          channels = c("C3-M2", "ECG"),
          events = events)

```

Once written on disk, MDF records can be read using the `read_mdf()` function. It quickly returns signals, events and metadata as a list.

```{r read_mdf}

mdf <- read_mdf("15012016HD")

```

## Sleep Stages

Hypnograms @AASMScoringManual can be plotted from stages data stored in a dataframe.

```{r plot_hypnogram}

plot_hypnogram(mdf$events)

```

## Epoching

```{r epoching}

reference <- hypnogram(mdf$events)
reference <- reference[-nrow(reference),]

epochs <- epochs(signals = lapply(mdf$channels,function(x){x$signal}),
                 sRates = lapply(mdf$channels,function(x){x$metadata$sRate}),
                 resample = 200,
                 epoch = reference,
                 startTime = as.numeric(as.POSIXct(mdf$metadata$startTime)))

```


## Electroencephalography

Fourier transforms are computed over EEG during sleep since 1942 @knottFourierTransformsElectroencephalogram1942 . Spectrograms of whole night signals can be plotted using the `spectrogram` function.

```{r spectrogram, warning=FALSE}

spectrogram(signal = mdf$channels$`C3-M2`$signal,
            sRate = mdf$channels$`C3-M2`$metadata$sRate,
            startTime = as.POSIXct(mdf$metadata$startTime))

```

### Spectral powers

```{r bands_computing}

bands <- lapply(epochs,function(x){
  apply(x, 2, function(y){
    bands_power(bands = list(c(0.5,3.5),c(3.5,7.5),c(7.5,13),c(13,30)),
                signal = y, sRate = 200,
                normalize = c(0.5,30))
  })
})

```

```{r bands_reshape}

c3m2 <- lapply(bands,function(x){
  unlist(x$`C3-M2`)
})
bands_df <- data.frame(matrix(unlist(c3m2), nrow=length(c3m2), byrow=TRUE))

colnames(bands_df) <- c("Delta","Theta","Alpha","Beta")
bands_df$stage <- reference$event
bands_df <- reshape2::melt(bands_df, id="stage")

summary(bands_df)
```

```{r bands_plot}

library(ggplot2)

pal <- c("#FF0000","#00A08A","#F98400","#5BBCD6")
ggplot(bands_df,aes(x=stage,y=value,fill=variable)) + 
  geom_boxplot() + theme_bw() +
  scale_fill_manual(values = pal) +
  theme(legend.title = element_blank()) +
  xlab("") + ylab("Normalized power") 

```

## Electrocardiography

`detect_rpeaks` implements the first part of the Pan & Tompkins algorithm @panRealTimeQRSDetection1985 to detect R peaks from an electrocardiogram (ECG) signal.

```{r detect_rpeaks}

library(ggplot2)

sRate <- 200

ecg <- data.frame(Volts = example_ecg_200hz,
                  Seconds = c(1:length(example_ecg_200hz))/sRate)

rpeaks <- detect_rpeaks(example_ecg_200hz, sRate)

ggplot(ecg,
       aes(x = Seconds,
           y = Volts)) +
  geom_line() + theme_bw() +
  geom_vline(data.frame(p = rpeaks),
             mapping = aes(xintercept = p),
             linetype="dashed",color = "red")

```

### Heart Rate Variability

Using the RHRV package @rodriguez-linaresOpenSourceTool2011, Heart Rate Variability (HRV) analysis can be conducted over whole records. 

Building and interpolating heart rate signal is the first step of an HRV analysis.

```{r hrv, message=FALSE, warning=FALSE}

library(RHRV)

peaks <- detect_rpeaks(signal = mdf$channels$ECG$signal,
                       sRate = mdf$channels$ECG$metadata$sRate)


hrv <- CreateHRVData()

hrv <- LoadBeatVector(HRVData = hrv,
                      beatPositions = peaks,
                      scale = 1,
                      datetime = format(as.POSIXct(mdf$metadata$startTime), "%d/%m/%Y %H:%M:%S"))

# Building the non-interpolated heart rate signal
hrv <- BuildNIHR(hrv)

# Filtering twice to eliminate all artifacts
hrv <- FilterNIHR(hrv)
hrv <- FilterNIHR(hrv)

# Interpolation of the instantaneous heart rate
hrv <- InterpolateNIHR(hrv)

PlotHR(hrv)

```

## Statistics computing

### Stages & scoring

`stages_stats` function computes various statistics from the hypnogram.

  * `rem_duration`: Total duration of REM sleep in minutes.
  * `n1_duration`: Total duration of N1 sleep in minutes.
  * `n2_duration`: Total duration of N2 sleep in minutes.
  * `n3_duration`: Total duration of N3 sleep in minutes.
  * `awa_duration`: Total duration of wake in minutes.
  * `tts`: Time To Sleep (N1+N2+N3+REM durations) in minutes.
  * `rem_tts`: REM over TTS duration ratio.
  * `n3_tts`: N3 over TTS duration ratio.
  * `n2_tts`: N2 over TTS duration ratio.
  * `n1_tts`: N1 over TTS duration ratio.
  * `tsp`: Total Sleep Period.
  * `sleep_efficiency`: Sleep Efficiency.
  * `sleep_latency`: Sleep Latency.
  * `rem_latency`: REM Sleep Latency.
  * `waso`: Wake After Sleep Onset.
  
```{r stages_stats}

stages_stats(example_hypnogram_30s)

```

## References

```{r cleanup, include=FALSE}

#unlink("15012016HD", recursive = TRUE)

```

