---
title: "Sleep Stages Classification Using Deep Learning"
author: "Paul Bouchequet"
output: rmarkdown::html_vignette
bibliography: Automatic_Stage_Classification.bibtex
vignette: >
  %\VignetteIndexEntry{automatic-stage-classification}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Do not run chunks if files are not present.
knitr::opts_chunk$set(
  eval = all(file.exists(
    "15012016HD.edf","15012016HD.csv",
    "SC4001E0-PSG.edf", "SC4002E0-PSG.edf",
    "SC4001EC-Hypnogram.edf", "SC4002EC-Hypnogram.edf")))

```

`rsleep` provides functions to score sleep stages from polysomnography data 
using deep neural networks. A pre-trained model can be downloaded and applied. 
As this model have been trained using polysomnography data from a single 
recorder( [Nox A1](https://noxmedical.com/products/nox-a1-psg-system/) designed 
by [Resmed](https://www.resmed.com/fr-fr/consumer.html) ), model can be trained 
on new data after generating a training database.

# Example data

First, download an example polysomnography record:

```{r download_edf, eval=FALSE}
download.file("https://osf.io/57j2u/download", "15012016HD.edf")
```

This polysomnography have been recorded at the Center for Sleep and Vigilance of
the Hôtel-Dieu, Paris, France, on a healthy subject. Only physiological signals 
are contained in this file. The corresponding hypnogram can be downloaded using 
using the following line:

```{r  download_hypnogram, eval=FALSE}
download.file("https://osf.io/h4ysj/download", "15012016HD.csv")
```

The hypnogram show sleep stages over time using consecutive 30 seconds epochs. 
Sleep experts usually score sleep records visually following the guidelines from
the American Association of Sleep Medicine (AASM) @AASMScoringManual. 
Five stages can be observed:

  * **AWA**: Awakeness.
  * **REM**: Rapid-Eye-Movement, REM (paradoxical) sleep, highlighted in red.
  * **N1**: Light sleep transitional stage.
  * **N2**: Light sleep.
  * **N3**: Slow-wave sleep.
  
This polysomnography has been visually scored by a sleep expert. Results of this
scoring can be found in the downloaded `.csv` file. This file has been exported 
from Noxturnal, the visual sleep scoring software published by Resmed. The 
`read_events_noxturnal()` function from the `rsleep` package can read this 
export and clean it to produce a standardized events dataframe. Then, the 
`plot_hypnogram()` function plots the hypnogram into a stage/time figure.

```{r, fig.width=7}
library(rsleep)

events <- read_events_noxturnal("15012016HD.csv")

plot_hypnogram(events)
```

# Applying a pre-trained model

Visual sleep scoring is an emprical science requiring a considerable amount of 
clinical knowledge @tongQuantitativeEEGAnalysis2009. However, many publications
showed scoring can be achieved by machine learning algorithms. In this field, 
deep neural networks showed the best results, similar to human sleep expert.

The `rsleep` package provides a pretrained model to score stages from raw sleep
signals. This model have been firstly trained on more than a hundred sleep 
records from the Hôtel-Dieu using the deep neural network described by Stanislas
Chamban & Al. in "A Deep Learning Architecture for Temporal Sleep Stage 
Classification Using Multivariate and Multimodal Time Series"
@chambonDeepLearningArchitecture2017.

To use this pre-trained model, call the `score_stages_edf()` function to score stages from the previously downloaded EDF file. `score_stages_edf()`, used with default parameters, downloads a pretrained model, preprocess the data, apply the model and returns the hypnodensity of the record. Hypnodensity shows the probabilities of each sleep stage by epoch. It has been introduced by Stephansen & Al. in "Neural network analysis of sleep stages enables efficient diagnosis of narcolepsy" @stephansenNeuralNetworkAnalysis2018.

```{r}
hypnodensity <- score_stages_edf("15012016HD.edf")
```

Returned hypnodensity can by plotted using the `plot_hypnodensity()` function.

```{r, fig.width=7}
plot_hypnodensity(hypnodensity)
```

The hypnogram can be computed form the hypnodensity taking the highest probability of each epoch.

```{r, fig.width=7}
plot_hypnogram(hypnodensity)
```

# Training a new model

Unfortunately, the weighted model provided by `rsleep` has only been trained on sleep records from a single polysomnograph. As it was roughly trained without extended preprocessing, trying to apply it to data from different recorders may result in false classifications. [Bit depth](https://etc.usf.edu/techease/win/images/what-is-bit-depth/), original sampling rate or [electrodes types](https://www.brainlatam.com/blog/wet-dry-active-and-passive-electrodes.-what-are-they-and-what-to-choose-413) can indeed vary between polysomnographs and produce different physiological signals.

The process to train a model using `rlseep` is pretty straightforward. First, the `generate_batches()` function creates batches files from EDF and events files. Then, the `train_batch()` trains the model using the previously generated batches. The model can be applied to new data using `score_stages_edf()` using the corresponding parameter.

```{r ml_workflow, echo=FALSE}
library(DiagrammeR)

create_graph() %>%
  add_node(label = "generate_batches()",
           node_aes = node_aes(fixedsize = FALSE,
                               shape = "rectangle")) %>%
  add_node(label = "train_batches()",
           node_aes = node_aes(fixedsize = FALSE,
                               shape = "rectangle")) %>%
  add_node(label = "score_stages_edf()",
           node_aes = node_aes(fixedsize = FALSE,
                               shape = "rectangle")) %>%
  add_edge(from = 1, to = 2) %>%
  add_edge(from = 2, to = 3) %>%
  render_graph(layout = "tree")
```

## Download example data

To illustrate the process of training a deep learning algorithm to classify sleep stages, let's download the first three records from the SleepEDFX extended database @kempAnalysisSleepdependentNeuronal2000 @goldbergerPhysioBankPhysioToolkitPhysioNet2000:

```{r sleepedfx_download}
records <- c("SC4001E0-PSG.edf", 
             "SC4002E0-PSG.edf",
             "SC4011E0-PSG.edf")

hypnograms <- c("SC4001EC-Hypnogram.edf", 
                "SC4002EC-Hypnogram.edf",
                "SC4011EH-Hypnogram.edf")

for(file in c(records,hypnograms)){
  if(!file.exists(file)){
    download.file(url = paste0(
      "https://physionet.org/files/sleep-edfx/1.0.0/sleep-cassette/",
      file, "?download"
    ), destfile = file)
  }
}
```

## Generating a training database

Six channels will be used to classify sleep stages:

```{r}
channels <- c(
  "EEG Fpz-Cz",
  "EEG Pz-Oz",
  "EOG horizontal",
  "EMG submental",
  "Temp rectal")
```

`write_batches_psg()` does not read events and stages in text files. Stages must be read from files previously and stored in a list:

```{r}
hypnograms <- lapply(hypnograms, function(x){
  read_events_sleepedfx(x)[-1,]
})
```

```{r}
write_batches_psg(
  records = records[1:2],
  events = hypnograms[1:2],
  batches_path = "./", 
  batches_size = 1000,
  channels = channels)
```

## Defining the model architecture

Sequential model implementation in Keras can be retreived using the `chambon2018()` @chambonDeepLearningArchitecture2017 function. Adding the parameter `weights` to `TRUE` downloads the weighted model. Number of channels input to the model is set using the `channels` parameters.

```{r}
model <- chambon2018(
  channels = length(channels))

summary(model)
```

## Training the model

`train_batches()` 

```{r}
batches <- list.files(pattern = "batch*")

model <- train_batches(
  model = model,
  batches = batches,
  epochs = 5)
```

## Applying the model

```{r}
hypnodensity <- score_stages_edf(
  edf = "SC4002E0-PSG.edf",
  model = keras::unserialize_model(model),
  channels = channels)

plot_hypnodensity(hypnodensity)

plot_hypnogram(
  read_events_sleepedfx("SC4002EC-Hypnogram.edf")[-1,])
```

```{r cleanup, include=FALSE}
file.remove(list.files(pattern = "*batch*"))
```

# References
