---
title: "Using rsleep and SleepCycles packages to detect sleep cycles"
output: 
  rmarkdown::html_vignette:
     toc: true
     toc_depth: 3
bibliography: Using_rsleep_and_SleepCycles_packages_to_detect_sleep_cycles.bibtex
vignette: >
  %\VignetteIndexEntry{Using rsleep and SleepCycles packages to detect sleep cycles}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r env, include = FALSE}

options(scipen=999)

options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

```

The [SleepCycles](https://CRAN.R-project.org/package=SleepCycles) package @BLUME2021101318 has been specifically developed to identify sleep cycles @Feinberg_Floyd_1979 and their corresponding NREM and REM components (known as (N)REM periods) from data that has been categorized based on AASM criteria for sleep staging @AASMScoringManual.

In the other hand, the [rsleep](https://rsleep.org/) package reads and analyze sleep data in various formats.

This vignette describes how to combine `SleepCycles` and `rsleep` packages to identify sleep cycles in sleep data and then leverage this material in sleep data analysis pipelines.

## Hypnogram

Sleep cycles can be identified from hypnograms. [`15012016HD.csv`](https://rsleep.org/data/15012016HD.csv) contains a hypnogram scored by a sleep expert using Noxturnal software published by ResMed. The rsleep package provides the `read_events_noxturnal()` function to read hypnograms in this format.

```{r hypnogram, echo=FALSE, message=FALSE, warning=FALSE}
library(rsleep)

if(!file.exists("15012016HD.csv")){
  download.file(
  url = "https://rsleep.org/data/15012016HD.csv",
  destfile = "15012016HD.csv",
  method="curl")}

events <- read_events_noxturnal("15012016HD.csv")

events = hypnogram(events)

plot_hypnogram(events)

```

## Formatting

Tne `SleepCycles` package only reads directories and files in specific format. Hypnograms must be converted in the appropriate arrangement before being written on disk in an explicit folder. The formatting function can be found in a forked version of the package, available on Github: `boupetch/SleepCycles`.

```{r formating, echo=TRUE, message=FALSE, warning=FALSE}

devtools::install_github("boupetch/SleepCycles")

dir = SleepCycles::format_events(events)

```

## Detection

The `SleepCycles()` function can now read the created directory and detect sleep cycles in the saved hypnograms.

```{r cycles, echo=TRUE, message=FALSE, warning=FALSE}

cycles = SleepCycles::SleepCycles(
  p = dir, 
  filetype = "txt", 
  plot = FALSE)

head(cycles)

```

```{r include=FALSE}

unlink(dir, recursive=TRUE)

```

Bound together, hypnogram and cycles dataframes make an enriched hypnogram dataset.

```{r}

hypnogram.full = cbind(events, cycles)

head(hypnogram.full)

```

## Plotting cycles

Compute the begining of each cyle to plot a dotted line at each cyle begining over the regular hypnogram plot.

```{r}

rsleep::plot_hypnogram_cycles(hypnogram.full)

```

## References
